Modelling the Retinotectal projection - can competition provide a stopping mechanism?

### 2022-03-03 paper/paper.tex
Working through the model description

We defined 4 receptor expression gradients arranged in orthogonal pairs
with the gradient of $r_0$ being orthogonal to that of $r_1$. $r_2$, whose
gradient is opposite to $r_0$, was orthogonal to $r_3$.

@@

Cells on the tectum express ligands, $L_i$, for the retinal receptors, also in
orthogonal pairs of gradients. Although several studies model tectal ligand
expression with exponential functions \citep{koulakov_stochastic_2004}, the
experimental evidence for the form of tectal ligand expression is more ambiguous than that for
retinal receptor expression (Fig.\,\ref{f:expressions} summarises results from
the literature). We kept in mind the
possibility that tectal ligand expression may be modelled by exponential,
linear or logarithmic functions, but initially, we set it to the same
exponential function used for retinal expression,
\begin{equation} \label{e:tecligexp}
L_i^{\text{exp}}(x,y) = 1.05 + 0.26 \exp(2.3 u),
\end{equation}
and used the substitutions $y$, $x$, $(1-y)$ or $(1-x)$ in place of $u$ to obtain
the ligand expression functions for $L_0$, $L_1$, $L_2$ and $L_3$
(Fig.\,\ref{f:ex}D).

@@

The form of the retinal receptor expression (Eq.\,\ref{e:retrcpt}) is based on
empirical measurement, but the tectal ligand expression
(Eq.\,\ref{e:tecligexp}) was chosen arbitrarily. To expand the pattern in
Fig.\,\ref{f:G}D so that it covers the entire tectum, we needed only to
adjust the tectal ligand expression, substituting the alternative exponential
expression:
%
\begin{equation} \label{e:tecligexp2}
L_i^{\text{exp}}(x,y) = 1.05 + 0.26 \exp(1.1 u)
\end{equation}
%
for Eq.\,\ref{e:tecligexp}.

@@

% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_wt_fig3.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_eE_G_wt_fig3_exit_true.png}
\caption{By tuning the tectal ligand expression (replacing Eq.\,\ref{e:tecligexp} with
 Eq.\,\ref{e:tecligexp2}), the pure gradient-following model can achieve a
 successful layout in which the axons are arranged evenly across the full
 tectum. At the centre of the tectum, where neither retinal receptor
 expression nor tectal ligand expression is high, the ordering is weaker than
 towards the edge. \textbf{A} Axon branches \textbf{B} axon
 centroids \textbf{C} Five selected axons.}
\label{f:G0}
\end{figure}

@@

The retinal ablation shown in Fig.\,\ref{f:Gsurg}D is also unsuccessful, with
the axons from the surviving nasal retina (the temporal half was removed)
failing to arrange themselves across the entire tectum, due to the pure
`position specification' which the dual gradient model achieves. Here, the
average axon position on the rostral-caudal axis of the tectum is plotted versus the average
nasal-temporal origin of the axon on the retina. Similarly,
the tectal ablation simulation fails to reproduce experimentally observed
results in which the retinal map `squashes up' within the remaining tectal
region. Instead, axons from the temporal retina arrange with normal spacing
and axons from the nasal retina are pushed to the caudal-most side of the
remaining tectum (Fig.\,\ref{f:Gsurg}E). A similar result occurs in the
mismatch experiment (Fig.\,\ref{f:Gsurg}F), but in this case there are no
axons from the temporal retina, which was been ablated and only the axons
from the nasal retina are present, arranged in a line.

@@

The model defined by the receptor and ligand expressions of Eqs.\,\ref{e:retrcpt}
\& \ref{e:tecligexp} along with Eq.\,\ref{e:mv2} has three
free parameters; $m_{\!_G}$, $m_c$ and $r_c$. To find the best combination of
these parameters, we reasoned that evolution would optimize the arrangement of
the wildtype pattern. We performed a simulated annealing optimization to
minimize the error metric, $\epsilon$, for the wildtype pattern
(Fig.\,\ref{f:G}E) at simulation time 1000. The parameters returned were
$m_{\!_G}$ = 0.003841, $m_c$ = 0.09959 and $r_c$ = 0.39918. Fig.\,\ref{f:GC}
shows the $\mathbf{GC}$ simulation with these parameters. Fish net plots at
$t=40$ and $t=1000$ show that the pattern develops quickly and compares well
with the expected layout (Fig.\,\ref{f:G}E). We investigated the response of
this model to the same set of surgical manipulations shown for the
chemoaffinity-only model in Fig.\,\ref{f:Gsurg}. We confirmed that the
addition of competition leads to a better correspondence to the experimental
results than for the chemoaffinity-only model (see supplementary
Fig.\,\ref{f:GCsurg}). We also confirmed the result
from \citet{simpson_simple_2011} that this description of a competitive
interaction is not able to reproduce the result of \citet{reber_relative_2004}
(Fig.\,\ref{f:GCreber}) Hmm. DID we? That graph looks nearly as good as the GJ
one...

@@

Because the competition effect, $\mathbf{C}_b$ (Eq.\,\ref{e:C}), does not
describe a signalling mechanism and cannot, on its own, explain all of the
experimental results, we looked for a receptor-ligand interaction that could
signal between axon branches and enable them to adjust their movements away
from one another.

@@

%We modelled interactions between axon growth cones due to signalling between
%receptors of the same type. For these interactions, repulsion caused a
%movement of branch $b$ along a unit vector, $\hat{\mathbf{x}}_{kb}$, from
%branch $k$ to branch $b$ causing them to move further apart; attraction caused
%the opposite movement. This interaction, $\mathbf{I}_b$, acting on branch $b$,
%is given by
%
\begin{equation}
\mathbf{I}_b
= \frac{1}{|B_{b,I}|} \sum_k^{nM} \hat{\mathbf{x}}_{kb}\,W_{\!_I} \qquad \mathrm{if}~r_{i,b}
/ r_{i,k} > 1.1
\end{equation}
%
$r_{i,k}$ is the expression of receptor $i$ on branch $k$ and the weight here
is defined as a function of the distance, $d_{kb}$, between branches:
%
\begin{equation} \label{e:WI}
W_{\!_I}(d_{kb}) = \begin{cases}
      1 - \frac{d_{kb}}{2r_{\!_I}}   & d_{kb} \leq 2r_{\!_I} \\
     0 & d_{kb} > 2r_{\!_I}
     \end{cases}
\end{equation}
%
$B_{b,I}$ is the set of growth cones within a distance $2r_{\!_I}$ of branch
$b$. $\hat{\mathbf{x}}_{kb}$ is as defined in Eq.\,\ref{e:C}.

@@

To avoid this asymmetry, we allow two receptor-ligand pairs, which are
expressed in opposing directions, to contribute to axon-axon interactions,
ensuring that interactions between single growth cones and clusters are not
dependent on their origin locations. Fig.\,\ref{f:expressions}C shows ligand
expressions representing an ephrin-A like expression ($l_0$) and an opposing
expression ($l_2$). It remains the case (with a suitable choice of threshold,
$s$) that branches originating from the same location on the temporal-nasal
axis will not experience competitive interactions.

Note that it would be possible to postulate two additional retinal ligand
expression fields (for $l_1$ and $l_3$), but we found this not to be
necessary; a single opposing pair of fields ensures that axons that originated
close together (and branches of the same axon) do not repel one another,
whereas axons that are close on the temporal-nasal axis, but far apart on the
dorso-ventral axis are generally driven far apart by chemotaxis, so whether
they compete or not when close together does not affect the behaviour of the
system.

% Now describe J model, and show the result.
We defined a movement vector arising from a multiplicative receptor-ligand
interaction, $\mathbf{J}_b$ (scaled by a parameter $m_{\!_J}$) to define the
movement vector, $\mathbf{M}_b$, for the $\mathbf{GJ}$ model as:
%
\begin{equation} \label{e:mv3}
 \mathbf{M}_{b} = m_{\!_G} \mathbf{G}_b + m_{\!_J} \mathbf{J}_b + 0.5 \mathbf{B}_b
\end{equation}
%
% noting that $m_J J_b$ \emph{replaces} $m_c C_b$.
%
$\mathbf{J}_b$ is defined as
%
\begin{equation}
\mathbf{J}_b = \frac{1}{|B_{b,J}|} \sum_k^{nM} \hat{\mathbf{x}}_{kb}\,W_{\!_J}\,Q_{\!_J}(d_{kb})
\end{equation}
%
where $W_{\!_J}$ is distance based weighting analogous to $W_{\!_I}$,
$B_{b,J}$ is analogous to $B_{b,I}$ and $Q_{\!_J}(d_{kb})$ is the signalling
strength between two growth cones of receptor-ligand interaction radius
$r_{\!_J}$ a distance $d_{kb}$ from one another, given by
%
\begin{equation}
Q_{\!_J}(d_{kb}) = \begin{cases}
                 F_i & \sum_i^N r_{i,b}\,l_{i,k} > s,~d_{kb} \leq 2r_{\!_J} \\
                 0 & \mathrm{otherwise}
     \end{cases}
\end{equation}
%
The sign of $Q_{\!_J}$ determines whether the interaction, $\mathbf{J}_b$, is
repulsive or attractive. $l_{i,k}$ is the expression of ligand $i$ on branch
$k$.

$\mathbf{J}_b$ would be equivalent to $\mathbf{I}_b$ if there was a single
$l_i$ expressed in the retina and its form was $1/r_i$. As well as allowing
for two, opp

With $m_{\!_G}$ fixed at the value 0.0012, and allowing the threshold
parameter $s$ to vary along with $m_{\!_J}$ and $r_{\!_J}$ we performed a
simulated annealing optimization. The parameters best able to recover the
wildtype pattern were $s=4.732$, $m_{\!_J} = 0.247$ and $r_{\!_J} =
0.079$. The simulation of the system with these parameters is shown in
Fig.\,\ref{f:GJ} and Fig.\,\ref{f:GJsurg}.

@@

\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_ee_GJ_best_1_reber.png}
\caption{$\mathbf{GJ}$ model, Reber et al.~manipulation. Look at THAT!}
\label{f:GJreber}
\end{figure}

>>>>

We defined 4 receptor expression gradients arranged in orthogonal pairs
with the gradient of $r_0$ being orthogonal to that of $r_1$. $r_2$, whose
gradient is opposite to $r_0$, was orthogonal to $r_3$ (Fig.\,\ref{f:expressions}B).

@@

Cells on the tectum express ligands, $L_i$, for the retinal receptors, also in
orthogonal pairs of gradients. Although several studies model tectal ligand
expression with exponential functions \citep{koulakov_stochastic_2004}, the
experimental evidence for the form of tectal ligand expression is more ambiguous than that for
retinal receptor expression (Fig.\,\ref{f:expressions} summarises results from
the literature). We kept in mind the
possibility that tectal ligand expression may be modelled by exponential,
linear or logarithmic functions, but initially, we set it to the same
exponential function used for retinal expression,
\begin{equation} \label{e:tecligexp}
L_i(x,y) = 1.05 + 0.26 \exp(2.3 u),
\end{equation}
and used the substitutions $y$, $x$, $(1-y)$ or $(1-x)$ in place of $u$ to obtain
the ligand expression functions for $L_0$, $L_1$, $L_2$ and $L_3$
(Fig.\,\ref{f:ex}D).

@@

The form of the retinal receptor expression (Eq.\,\ref{e:retrcpt}) is based on
empirical measurement, but the tectal ligand expression
(Eq.\,\ref{e:tecligexp}) was chosen arbitrarily. To expand the pattern in
Fig.\,\ref{f:G}D so that it covers the entire tectum, we needed only to
adjust the tectal ligand expression, substituting the alternative exponential
expression:
%
\begin{equation} \label{e:tecligexp2}
L_i(x,y) = 1.05 + 0.26 \exp(1.1 u)
\end{equation}
%
for Eq.\,\ref{e:tecligexp}.

@@

% ./build/sim/agent/agent1 configs/simpler/m_eE_G.json  configs/simpler/e_wt_fig3.json
\begin{figure}
\includegraphics[width=\linewidth]{./images/j4_eE_G_wt_fig3_exit_true.png}
\caption{By tuning the tectal ligand expression (replacing Eq.\,\ref{e:tecligexp} with
 Eq.\,\ref{e:tecligexp2}), the pure gradient-following model can achieve a
 successful layout in which the axons are arranged evenly across the full
 tectum. At the centre of the tectum, where neither retinal receptor
 expression nor tectal ligand expression is high, the ordering is weaker than
 towards the edge. \textbf{A} Axon branches \textbf{B} axon
 centroids \textbf{C} The pattern metric $\eta$ is zero, as it was in
 Fig.\,\ref{f:G}F and $\epsilon$ approaches a much lower value as $t \to \infty.}
\label{f:G0}
\end{figure}

@@

The retinal ablation shown in Fig.\,\ref{f:Gsurg}D is also unsuccessful, with
the axons from the surviving nasal retina (the temporal half was removed)
failing to arrange themselves across the entire tectum, due to the pure
`position specification' which the dual gradient model achieves. Here, the
average axon position on the rostral-caudal axis of the tectum is plotted versus the average
nasal-temporal origin of the axon on the retina. Similarly,
the tectal ablation simulation fails to reproduce experimentally observed
results in which the retinal map `squashes up' within the remaining tectal
region. Instead, axons from the temporal retina arrange with normal spacing
and axons from the nasal retina are pushed to the caudal-most side of the
remaining tectum (Fig.\,\ref{f:Gsurg}E). A similar result occurs in the
mismatch experiment (Fig.\,\ref{f:Gsurg}F), but in this case there are no
axons from the temporal retina, which was ablated and only the axons
from the nasal retina are present, arranged in a line.

@@

The model defined by the receptor and ligand expressions of Eqs.\,\ref{e:retrcpt}
\& \ref{e:tecligexp} along with Eq.\,\ref{e:mv2} has three
free parameters; $m_{\!_G}$, $m_c$ and $r_c$. To find the best combination of
these parameters, we reasoned that evolution would optimize the arrangement of
the wildtype pattern. We performed a simulated annealing optimization to
minimize the error metric, $\epsilon$, for the wildtype pattern
(Fig.\,\ref{f:G}E) at simulation time 1000. The parameters returned were
$m_{\!_G}$ = 0.003841, $m_c$ = 0.09959 and $r_c$ = 0.39918. Fig.\,\ref{f:GC}
shows the $\mathbf{GC}$ simulation with these parameters. Fish net plots at
$t=40$ and $t=1000$ show that the pattern develops quickly and compares well
with the expected layout (Fig.\,\ref{f:G}E). We investigated the response of
this model to the same set of surgical manipulations shown for the
chemoaffinity-only model in Fig.\,\ref{f:Gsurg}. We confirmed that the
addition of competition leads to a better correspondence to the experimental
results than for the chemoaffinity-only model (see supplementary
Fig.\,\ref{f:GCsurg}).

@@

Because the competition effect, $\mathbf{C}_b$ (Eq.\,\ref{e:C}), does not
describe a signalling mechanism, and because \citet{simpson_simple_2011}
reported that an additional mechanism is required to explain the gene
manipulation experiments of \citet{brown_topographic_2000}
and \citet{reber_relative_2004}, we looked for a receptor-ligand interaction
that could signal between axon branches and enable them to adjust their
movements away from one another.

@@

%We modelled interactions between axon growth cones due to signalling between
%receptors of the same type. For these interactions, repulsion caused a
%movement of branch $b$ along a unit vector, $\hat{\mathbf{x}}_{kb}$, from
%branch $k$ to branch $b$ causing them to move further apart; attraction caused
%the opposite movement. This interaction, $\mathbf{I}_b$, acting on branch $b$,
%is given by
%
\begin{equation}
\mathbf{I}_b
= \frac{1}{|B_{b,I}|} \sum_k^{nM} \hat{\mathbf{x}}_{kb}\,W_{\!_I} \qquad \mathrm{if}~r_{i,b}
/ r_{i,k} > 1.1
\end{equation}
%
$r_{i,k}$ is the expression of receptor $i$ on branch $k$ and the weight here
is defined as a function of the distance, $d_{kb}$, between branches:
%
\begin{equation} \label{e:WI}
W_{\!_I}(d_{kb}) = \begin{cases}
      1 - \frac{d_{kb}}{2r_{\!_I}}   & d_{kb} \leq 2r_{\!_I} \\
     0 & d_{kb} > 2r_{\!_I}
     \end{cases}
\end{equation}
%
$B_{b,I}$ is the set of growth cones within a distance $2r_{\!_I}$ of branch
$b$. $\hat{\mathbf{x}}_{kb}$ is as defined in Eq.\,\ref{e:C}. This mechanism
adds a weighted repulsion if the ratio of receptor expression on
branch $b$ and the expression on branch $k$ exceeds the threshold of 1.1.

@@

To avoid this asymmetry, we allowed two receptor-ligand pairs, which are
expressed in opposing directions, to contribute to axon-axon interactions,
ensuring that interactions between single growth cones and clusters are not
dependent on their origin locations. Fig.\,\ref{f:ex}C shows ligand
expressions representing an ephrin-A like expression ($l_0$) and an opposing
expression ($l_2$). We adopted a form for the ligand expression such that
$r_i \times l_i = 4.7727$, with the scalar constant chosen so that the retinal
ligand expression, $l_i(x,y)$ has approximately the same start and end values
as the exponential retinal receptor expression:
%
\begin{equation}
 \label{e:retligexp}
l_i(x,y) = \frac{4.7727}{1.05 + 0.26 \exp (2.3 u)}
\end{equation}
$u$ was substituted with $(1-x)$ for $l_0$ or $x$ for $l_2$ (Fig.\,\ref{f:ex}C).
%
%With a suitable choice of threshold, $s$, branches originating from the same
%location on the temporal-nasal axis will not experience competitive
%interactions.

Note that it would be possible to postulate two additional retinal ligand
expression fields (for $l_1$ and $l_3$), but we found this not to be
necessary; a single opposing pair of fields can ensure that most pairs of
axons are able to signal the necessity for a competitive interaction. We set
$l_1 = l_3 = 0$.

% Now describe J model, and show the result.
We defined a movement vector arising from a multiplicative receptor-ligand
interaction, $\mathbf{J}_b$ (scaled by a parameter $m_{\!_J}$) to define the
movement vector, $\mathbf{M}_b$, for the `$\mathbf{GJ}$ model' as:
%
\begin{equation} \label{e:mv3}
 \mathbf{M}_{b} = m_{\!_G} \mathbf{G}_b + m_{\!_J} \mathbf{J}_b + 0.5 \mathbf{B}_b
\end{equation}
%
% noting that $m_J J_b$ \emph{replaces} $m_c C_b$.
%
$\mathbf{J}_b$ is defined as
%
\begin{equation}
\mathbf{J}_b = \frac{1}{|B_{b,J}|} \sum_k^{nM} \hat{\mathbf{x}}_{kb}\,W_{\!_J}\,Q_{\!_J}(d_{kb}, \mathbf{r}_{b}, \mathbf{l}_{k})
\end{equation}
%
where $W_{\!_J}$ is distance based weighting analogous to $W_{\!_I}$ and
$B_{b,J}$ is analogous to $B_{b,I}$. $Q_{\!_J}$ determines if the signals
between a branch $b$ expressing receptors $\mathbf{r}_b$ and branch $k$
expressing ligands $\mathbf{l}_k$, separated by a distance $d_{kb}$ will cause
$b$ to move away from $k$. $Q_{\!_J}(d_{kb}, \mathbf{r}_{b}, \mathbf{l}_{k})$
can be expressed as
%
\begin{equation}
Q_{\!_J}(d_{kb}, \mathbf{r}_{b}, \mathbf{l}_{k}) = \begin{cases}
                 0 & \mathrm{if}~-F_i\,r_{i,b}\,l_{i,k} <
                 s,\,\forall{i}~\mathrm{or}~d_{kb} > 2r_{\!_J} \\
                 1 & \mathrm{otherwise}
     \end{cases}
\end{equation}

where $r_{\!_J}$ is the receptor-ligand interaction radius, $l_{i,k}$ (the
expression of ligand $i$ on branch $k$) is the $i^{\mathrm{th}}$ element of
$\mathbf{l}_k$ and $r_{i,b}$ is the $i^{\mathrm{th}}$ element of
$\mathbf{r}_b$. $F_i$, defined in Eq.\,\ref{e:G}, is -1 if the $r_{i}\,l_{i}$
interaction signals repulsion and +1 if it signals attraction. $s$ is a
positive signal threshold whose value should be around 4.7727; if higher, then
branches whose retinal origin is close will not experience a competitive
interaction. If $s\leq 4.7727$, then \emph{any} pair of branches will compete as they
approch each other. With the constraint $s>0$ comes the assumption that
attractive interactions will have no effect on the competitive movement
vector---a receptor-ligand pair that signals attraction cannot cause
$Q_{\!_J}$ to take the value 1. $Q_{\!_J}$ is set to 1 only if at least one
repulsive signal exceeds the threshold, $s$ and the distance from branch
$b$ to branch $k$ is smaller than $2 r_{\!_J}$.

Note that $\mathbf{J}_b$ would be equivalent to $\mathbf{I}_b$ (with $s=1.1$) if there
was a single $l_i$ expressed in the retina and its form was $1/r_i$.

% Now the simulation.
Adopting retinal expressions given in Eqs.\,\ref{e:retrcpt}
and \ref{e:retligexp}, we performed a simulated annealing optimization to find
the best parameters for the $\mathbf{GJ}$ model.  With $m_{\!_G}$ fixed at the
value 0.0012, we allowed the threshold parameter $s$ to vary along with
$m_{\!_J}$ and $r_{\!_J}$ . The parameters best able to recover the wildtype
pattern were $s=4.732$, $m_{\!_J} = 0.247$ and $r_{\!_J} = 0.079$. Note that
the discovered value for $s$ is less than 4.7727 meaning that for \emph{any
pair of branches}, either $r_0\times l_0$ or $r_2\times l_2$ will exceed $s$.

The
simulation of the system with these parameters is shown in Fig.\,\ref{f:GJ}
and Fig.\,\ref{f:GJsurg}.

@@

\begin{figure}
\includegraphics[width=\linewidth]{./images/EphA_manipulations.png}
\caption{$\mathbf{GJ}$ model, Brown et al./Reber et al.~manipulations.}
\label{f:GJreber}
\end{figure}
